Computing {\em the least commitment} action model from observations of plan executions

### 2018-11-26 doc/ijcai-19-conformant/conformant-learning.tex
added conformant planning model

\title{Computing the least commitment action models from observations of plan executions}

@@

\section{Introduction}

\section{Background}
This section formalizes the models for {\em classical planning} and for the {\em observation} of the execution of a classical plan.
\subsection{Classical planning with conditional effects}
$F$ is the set of {\em fluents} or {\em state variables} (propositional variables). A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e. either~$l=f$ or $l=\neg f$. $L$ is a set of literals that represents a partial assignment of values to fluents, and $\mathcal{L}(F)$ is the set of all literals sets on $F$, i.e.~all partial assignments of values to fluents. A {\em state} $s$ is a full assignment of values to fluents. We explicitly include negative literals $\neg f$ in states and so $|s|=|F|$ and the size of the state space is $2^{|F|}$.

@@

A {\em planning problem} is defined as a tuple $P=\tup{F,A,I,G}$, where $I$ is the initial state in which all the fluents of $F$ are assigned a value true/false and $G$ is the goal set. A {\em plan} $\pi$ for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$, and $|\pi|=n$ denotes its {\em plan length}. The execution of $\pi$ in the initial state $I$ of $P$ induces a {\em trajectory} $\tau(\pi,P)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, it holds $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$. A trajectory $\tau(\pi,P)$ that solves $P$ is one in which $G \subseteq s_n$.

@@

\subsection{The observation model}
Given a planning problem $P=\tup{F,A,I,G}$, a plan $\pi$ and a trajectory $\tau(\pi,P)$, we define the \emph{observation of the trajectory} as an interleaved combination of actions and states that represents the observation from the execution of $\pi$ in $P$. Formally, $\mathcal{O}(\tau)=\tup{s_0^o,a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, $s_0^o=I$, and:

%which indicates that we observe $l$ actions, $1\leq l\leq |\pi|$, and $m$ states, $1\leq m \leq |\pi|+1$, from $\tau(\pi,P)$:

\begin{itemize}
\item The {\bf observed actions} are consistent with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. Specifically, the number of observed actions, $l$, can range from $0$ (fully unobservable action sequence) to $|\pi|$ (fully observable action sequence).
\item The {\bf observed states} $\tup{s_0^o, s_1^o, \ldots, s_m^o}$ is a sequence of possibly {\em partially observable states}, except for the initial state $s_0^o$, which is fully observable. A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,P)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states, $m$, range from 1 (the initial state, at least), to $|\pi|+1$, and the observed intermediate states will comprise a number of fluents between $[1,|F|]$.
    %Exceptionally, $s_m^o$ cannot be fully unobservable for the purpose of our task (we will elaborate on this issue later on).
\end{itemize}

@@

\subsection{Conformant planning}

\section{Learning the least commitment action models from observations}

@@

\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$.

\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in a {\em blocksworld} one can claim that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in any of these lists because of the semantic of the {\tt\small on} predicate that codes which block is in top of other block. 
\end{itemize}

@@

\subsection{A propositional encoding for \strips\ action schema}
Given a \strips\ action schema $\xi$, a propositional encoding for the {\em preconditions}, {\em negative} and {\em positive} effects of that schema can be represented with fluents of the kind $[pre|del|add]\_e\_\xi$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of interpretations of predicates $\Psi$ over the corresponding variable names $\Omega_\xi$. Figure~\ref{fig:encoding} shows the propositional encoding for the six action schema defined in Figure~\ref{fig:model-example}.

@@

\subsection{Learning the least commitment models with conformant planning}

\section{Evaluation}
\section{Conclusions}

>>>>

\title{Computing {\em the least commitment} action model from observations of plan executions}

@@

\section{Introduction}
Given the partial observation of a plan execution, this paper formalizes the task of computing the exact set of action models that are {\em conformant} with the given observation. This task is of interest because it allows to scallably learn action models from arbitrary large sets of observations generated with that models. The approach we follow for addressing this learning task is compiling it into {\em conformant planning}.

\section{Background}
This section formalizes the {\em classical} and the {\em conformant} planning models as well as the {\em observation} model we follow to observe the execution of sequential plans.
\subsection{Classical planning with conditional effects}
$F$ is the set of {\em fluents} or {\em state variables} (propositional variables). A {\em literal} $l$ is a valuation of a fluent $f\in F$, i.e. either~$l=f$ or $l=\neg f$. $L$ is a set of literals that represents a partial assignment of values to fluents, and $\mathcal{L}(F)$ is the set of all literals sets on $F$, i.e.~all partial assignments of values to fluents. A {\em state} $s$ is a full assignment of values to fluents. We explicitly include negative literals $\neg f$ in states and so $|s|=|F|$ and the size of the state space is $2^{|F|}$.

@@

A {\em planning problem} is defined as a tuple $P=\tup{F,A,I,G}$, where $I$ is the initial state in which all the fluents of $F$ are assigned a value {\em true\textbar false} and $G$ is the set of goal conditions over the state variables in $F$. A {\em plan} $\pi$ for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$, and $|\pi|=n$ denotes its {\em plan length}. The execution of $\pi$ in the initial state $I$ of $P$ induces a {\em trajectory} $\tau(\pi,s_0)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, it holds $\rho(s_{i-1},a_i)$ and $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ solves $P$ iff the induced {\em trajectory} $\tau(\pi,s_0)$ reaches a final state $G \subseteq s_n$.

@@

\subsection{The observation model}
Given a planning problem $P=\tup{F,A,I,G}$, a plan $\pi$ and a trajectory $\tau(\pi,s_0)$, we define the \emph{observation of the trajectory} as an interleaved combination of actions and states that represents the observation from the execution of $\pi$ in $P$. Formally, $\mathcal{O}(\tau)=\tup{s_0^o,a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, $s_0^o=I$, and:
\begin{itemize}
\item The {\bf observed actions} are consistent with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. Specifically, the number of observed actions, $l$, can range from $0$ (fully unobservable action sequence) to $|\pi|$ (fully observable action sequence).
\item The {\bf observed states} $\tup{s_0^o, s_1^o, \ldots, s_m^o}$ is a sequence of possibly {\em partially observable states}, except for the initial state $s_0^o$, which is fully observable. A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Whatever the sequence of observed states of $\mathcal{O}(\tau)$ is, it must be consistent with the sequence of states of $\tau(\pi,s_0)$, meaning that $\forall i, s_i^o \subseteq s_i$. In practice, the number of observed states, $m$, range from 1 (the initial state, at least), to $|\pi|+1$, and the observed intermediate states will comprise a number of fluents between $[1,|F|]$.
    %Exceptionally, $s_m^o$ cannot be fully unobservable for the purpose of our task (we will elaborate on this issue later on).
\end{itemize}

@@

\subsection{Conformant planning}
{\em Conformant planning} refers to planning with incomplete information about the initial situation, no sensing, and where goals are to be achieved with certainty in spite of the uncertainty in the initial situation~\cite{smith1998conformant,goldman1996expressive}.

Syntactically, conformant planning problems are expressed in compact form through a set of state variables. A {\em conformant planning problem} is defined as a tuple $P_c=\tup{F,A,\Upsilon,G}$, where $\Upsilon$ is a set of clauses over literals $l=f$ or $l=\neg f$ (for $f\in F$) that define the set of possible initial states.

A solution to a conformant planning problem is an action sequence that maps each possible initial state into a goal state. More precisely an action sequence $\pi=\tup{a_1, \ldots, a_n}$ is a {\em conformant plan} for $P_c$ iff for each possible {\em trajectory} $\tau(\pi,s_0)=\tup{s_0, a_1, s_1, \ldots, a_n, s_n}$ such that $s_0$ is a valuation that satisfies $\Upsilon$ and it holds that the trajectory reaches $G \subseteq s_n$.

\section{Learning {\em the least commitment} action model from observations}

@@

\begin{enumerate}
\item {\em Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$.

\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance, in a {\em blocksworld} one can claim that {\small\tt on($v_1$,$v_1$)} and {\small\tt on($v_2$,$v_2$)} will not appear in any of these lists because of the semantic of the {\tt\small on} predicate that codes which block is in top of other block. 
\end{enumerate}

@@

\begin{enumerate}
\item[3.] {\em Observation constraints}. An observation of the execution of a plan further constraints the space of possible action schemata. This \emph{semantic knowledge} included in the observations introduce a third type of constraints, that we will call {\em observation constraints}, and that can be added to the set $\mathcal{C}$.
\end{enumerate}

\subsection{Learning the least commitment models with conformant planning}
Given a \strips\ action schema $\xi$, a propositional encoding for the {\em preconditions}, {\em negative} and {\em positive} effects of that schema can be represented with fluents of the kind $[pre|del|add]\_e\_\xi$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of interpretations of predicates $\Psi$ over the corresponding variable names $\Omega_\xi$. Figure~\ref{fig:encoding} shows the propositional encoding for the six action schema defined in Figure~\ref{fig:model-example}.

@@

\section{Evaluation}

\section{Conclusions}
